---
title: "Reproducible Research -- Course Project 1"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading and Preprocessing the Data
1. Load required libraries
```{r setup1, message=FALSE, warning=FALSE}
library(dplyr)
library(dtplyr)
library(data.table)
library(ggplot2)
```

2. Read the data from the supplied .csv file
```{r read}
# Read the data
setwd("C:/home/Data-Science-Course/Reproducible-Research/Course-Project-1")
activity <- read.csv("activity.csv")
```
3. Do a little preprocessing of this data
```{r}
# Convert the data frame to a data.table
activity <- tbl_dt(activity)
# Convert the date variable from character to Date
activity$date <- as.Date(as.character(activity$date))
```
# What is mean total number of steps taken per day?
1. Calculate the total number of steps per day
```{r}
# (Ignore NA values)
total_steps_per_day <- activity %>% na.omit() %>% group_by(date) %>% summarise(.,sum=sum(steps))
```
2. Plot the result as a histogram
```{r first_histogram}
ggplot(data=total_steps_per_day,aes(sum)) + geom_histogram(binwidth=500,col="blue",fill="green") + scale_x_continuous(breaks=seq(0,max(total_steps_per_day$sum),by=2500)) + labs(title="Histogram of Total Number of Steps Per Day",x="Total Steps Per Day")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r mean_and_median0}
summary(total_steps_per_day$sum)
```

# What is the average daily activity pattern?
1. Create a function "hackInterval()" that will generate nice x-axis labels representing 24-hour (a.k.a. "metric") time
```{r hack_interval}
hackInterval <- function(interval) {
    interval <- as.character(interval)
    num_chars <- nchar(interval)
    if(num_chars==1) {
        mod_interval <- "00h0"
        mod_interval <- paste(mod_interval,interval,sep="")
    }
    else if(num_chars==2) {
        mod_interval <- "00h"
        mod_interval <- paste(mod_interval,interval,sep="")
    }
    else if(num_chars==3) {
        split_interval <- strsplit(interval,'(?=^\\d{1}(?=\\d{2,}))',perl=T)[[1]]
        leading <- split_interval[1]
        trailing <- split_interval[2]
        mod_interval <- paste("0",leading,sep="")
        mod_interval <- paste(mod_interval,"h",sep="")
        mod_interval <- paste(mod_interval,trailing,sep="")
    }
    else if(num_chars==4) {
        lead_pattern <- "(^\\d{2})"
        trail_pattern <- "(\\d{2}$)"
        m_lead <- regexec(lead_pattern,interval)
        leading <- regmatches(interval,m_lead)[[1]][1]
        m_trail <- regexec(trail_pattern,interval)
        trailing <- regmatches(interval,m_trail)[[1]][1]
        mod_interval <- paste(leading,"h",sep="")
        mod_interval <- paste(mod_interval,trailing,sep="")
    }
    mod_interval
}
```
2. Create a data frame with the mean number of steps per 5-minute interval calculated (uses the hackInterval() function defined above)
```{r data_table_for_plotting}
activity_for_plotting2 <- activity %>% group_by(interval) %>% na.omit() %>% select(steps,date,interval) %>% summarise(mean_num_steps=mean(steps)) %>% rowwise() %>% mutate(mod_interval=hackInterval(interval))
```
3. Plot the mean daily activity pattern (ignoring NAs) as a time series
```{r time_series_plot_of_mean_num_steps}
ggplot(data=activity_for_plotting2,aes(interval,mean_num_steps)) + geom_line(color="darkgreen",size=1) + xlab("5-min Intervals Starting at Midnight") + ylab("Mean Number of Steps During Interval") + scale_x_continuous(breaks=activity_for_plotting2$interval[seq(1,length(activity_for_plotting2$interval),12)],labels=activity_for_plotting2$mod_interval[seq(1,length(activity_for_plotting2$mod_interval),12)]) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Average Daily Activity Pattern")
```

4. Answer the question: "Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?"
- Use dplyr's slice() function:
```{r slice}
activity_for_plotting2 %>% slice(which.max(mean_num_steps))
```

This result shows that the 5-minute interval beginning at 8:35 AM, on average across all of the days in the dataset, has the highest number of steps taken.

# Imputing missing values
1. Define a custom function to count NAs
```{r}
count_na <- function(x) sum(is.na(x))
```

2. Now use R's apply() function together with dplyr's mutate() function, to count the NAs in the original dataset:
```{r}
activity_with_na_sum <- activity %>% mutate(count_na=apply(.,1,count_na)) %>% summarise(.,total_num_NAs=sum(count_na))
```

Therefore, the total number of NAs in the dataset is 2304.

3. Devise a strategy for filling in all of the missing values in the dataset. (I used the mean, calculated above for each 5-minute interval over the entire measurement period, to fill in any NAs during that 5-minute interval.)
```{r fill_NAs}
# Save the original dataframe
activity_mod <- activity
# Add a new variable to the original data, which is the
# mean number of steps for each 5-minute interval during
# the measurement period. This was calculated earlier and # is the mean_num_steps variable in the 
# activity_for_plotting2 data frame/table
activity_mod$mean_num_steps <- activity_for_plotting2$mean_num_steps
```

4. Create a new dataset that is equal to the original dataset but with the missing data filled in
```{r NAs_filled}
# Where there is an NA value in the steps variable,
# replace it with the mean value for that 5-minute
# interval 
activity_mod_w_nas_filled <- activity_mod %>% mutate(steps=ifelse(is.na(steps),mean_num_steps,steps))
# Delete the mean_num_steps variable from the dataframe
# since its values have now been copied
activity_mod_w_nas_filled <- activity_mod_w_nas_filled %>% select(-mean_num_steps)
```
5. Plot the histogram of the data
```{r hist_steps}
# Transform the dataframe/table to a form required for
# plotting
total_steps_per_day_w_imputed_values <- activity_mod_w_nas_filled %>% na.omit() %>% group_by(date) %>% summarise(sum=sum(steps))
# Create the histogram
ggplot(data=total_steps_per_day_w_imputed_values,aes(sum)) + geom_histogram(binwidth=500,col="blue",fill="green") + scale_x_continuous(breaks=seq(0,max(total_steps_per_day_w_imputed_values$sum),by=2500)) + labs(title="Histogram of Total Number of Steps Per Day (with imputed values)",x="Total Steps Per Day")
```

Here are the summary statistics for the total number of steps taken per day (after NA values are replaced by imputed values):
```{r summary}
summary(total_steps_per_day_w_imputed_values$sum)
```

Recall that the summary statistics for the total number of steps taken per day (with NA values ignored) was:
```{r summary_raw}
summary(total_steps_per_day$sum)
```

Therefore we can conclude that there is an effect on the summary statistics due to imputing the NA values, and this effect is to increase the first and third quartile values, but the mean, min and max are not affected.

# Are there differences in activity patterns between weekdays and weekends?
1. Add a "day of week" variable to the dataframe containing imputed values for NAs, using the weekdays() function as suggested
```{r use_weekdays}
activity_mod_w_nas_filled$day_of_week <- weekdays(activity_mod_w_nas_filled$date,abbr=T)

```
2. Add a factor variable "weekday or weekend" to the same dataframe, which is based on the day of the week calculated above
```{r add_wknd}
activity_mod_w_nas_filled_w <- activity_mod_w_nas_filled %>% select(steps,date,interval,day_of_week) %>% mutate(weekday_or_weekend=ifelse(grepl("S(at|un)",day_of_week),"Weekend","Weekday"))
```

3. Convert the variable to factor class
```{r cvt_factor}
activity_mod_w_nas_filled_w$weekday_or_weekend <- factor(activity_mod_w_nas_filled_w$weekday_or_weekend)
```

4. Transform the dataframe into a form for plotting. Grouping by weekday/weekend and interval is the key here.
```{r transform_for_plotting}
weekday_or_weekend_5_min_avg_activity <- activity_mod_w_nas_filled_w %>% group_by(weekday_or_weekend,interval) %>% na.omit() %>% select(steps,date,interval) %>% summarise(mean_num_steps=mean(steps)) %>% rowwise() %>% mutate(mod_interval=hackInterval(interval))
```

5. Create the panel plot, one for weekday activity and the other for weekend activity
```{r panel_plot}
ggplot(data=weekday_or_weekend_5_min_avg_activity,aes(interval,mean_num_steps)) + geom_line(color="darkblue",size=1) + xlab("5-min Intervals Starting at Midnight") + facet_grid(. ~ weekday_or_weekend) + ylab("Mean Number of Steps During Interval") + scale_x_continuous(breaks=activity_for_plotting2$interval[seq(1,length(activity_for_plotting2$interval),12)],labels=activity_for_plotting2$mod_interval[seq(1,length(activity_for_plotting2$mod_interval),12)]) + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + ggtitle("Average Daily Activity Pattern")
```
